A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-07-31 manuscript.qmd
set preprint version to 3.5 (between major psyarxiv releases)
added a comment on how sensitivity analysis can be used to investigate the consequences of a mismatch between data-generating process and statistical model

    data-sharing: This is version 3 of our preprint published at <https://doi.org/10.31234/osf.io/rpjem>. All materials (reproducible manuscript, including R code and all simulation results) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/dhwf4/>. A Quarto Manuscripts website is hosted at <https://Timo-Ko.github.io/glmm_simulation_tutorial/>.

@@

In our case study, we have performed simulation-based sample size planning from a single set of parameter values that reflect our assumptions of an expected effect size. Instead of extracting this expected effect size from meta-analyses or pilot data, which has been the main focus of previous tutorials [e.g., @kumleEstimatingPowerGeneralized2021], we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge. When sample sizes are chosen based on the results of our simulation-based power analysis, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present (or estimate the effect with satisfying precision). However, if the true effect is indeed smaller, power (or precision) will be lower, and the study might not be sufficiently informative. A common, more conservative strategy for sample size justification is to perform sample size planning for the smallest effect size of interest (SESOI). An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero [@kingPointMinimalImportant2011; @lakensEquivalenceTestingPsychological2018]. Unfortunately, specifying a plausible SESOI is a challenging task (for strategies on how to specify a SESOI, see @lakensSampleSizeJustification2022, @riesthuisSimulationBasedPowerAnalyses2024, or @lakensImprovingYourStatistical2022). When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power or precision for different effect sizes in a *sensitivity power analysis* [@lakensSampleSizeJustification2022]. By simulating power (or precision) for different effect sizes (in addition to the different number of subjects and items), one can make sure that power (or precision) would still be sufficient to detect smaller effect sizes than our expected effect or at least get an impression of how strongly power (or precision) depends on the size of the true effect. For our case study that investigates combined hypotheses in a GLMM modeling framework, sensitivity analysis would require manually specifying additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question. Power (or precision) could then be simulated for several of these scenarios (across different numbers of subjects and items, as considered earlier). Although recommended, we do not present a sensitivity analysis for our case study to keep the length of the tutorial manageable.

>>>>

    data-sharing: This is version 3.5 of our preprint published at <https://doi.org/10.31234/osf.io/rpjem>. All materials (reproducible manuscript, including R code and all simulation results) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/dhwf4/>. A Quarto Manuscripts website is hosted at <https://Timo-Ko.github.io/glmm_simulation_tutorial/>.

@@

In our case study, we have performed simulation-based sample size planning from a single set of parameter values that reflect our assumptions of an expected effect size. Instead of extracting this expected effect size from meta-analyses or pilot data, which has been the main focus of previous tutorials [e.g., @kumleEstimatingPowerGeneralized2021], we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge. When sample sizes are chosen based on the results of our simulation-based power analysis, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present (or estimate the effect with satisfying precision). However, if the true effect is indeed smaller, power (or precision) will be lower, and the study might not be sufficiently informative. A common, more conservative strategy for sample size justification is to perform sample size planning for the smallest effect size of interest (SESOI). An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero [@kingPointMinimalImportant2011; @lakensEquivalenceTestingPsychological2018]. Unfortunately, specifying a plausible SESOI is a challenging task (for strategies on how to specify a SESOI, see @lakensSampleSizeJustification2022, @riesthuisSimulationBasedPowerAnalyses2024, or @lakensImprovingYourStatistical2022). When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power or precision for different effect sizes in a *sensitivity power analysis* [@lakensSampleSizeJustification2022]. By simulating power (or precision) for different effect sizes (in addition to the different number of subjects and items), one can make sure that power (or precision) would still be sufficient to detect smaller effect sizes than our expected effect or at least get an impression of how strongly power (or precision) depends on the size of the true effect. For our case study that investigates combined hypotheses in a GLMM modeling framework, sensitivity analysis would require manually specifying additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question. Power (or precision) could then be simulated for several of these scenarios (across different numbers of subjects and items, as considered earlier). In steps 2 and 4 of the tutorial, we have briefly discussed scenarios where the applied statistical model is more (or less) parsimonious than the data-generating process. Sensitivity analysis can be used to assess the consequences of a mismatch by investigating different combinations of data-generating processes and statistical models. Such extended simulations can be a great resource to make an informed decision about the concrete model to estimate for the planned study where the true data-generating process is unknown. Although recommended, we do not present a sensitivity analysis for our case study to keep the length of the tutorial manageable.
